Models, derivatives, response diversity

### 2024-07-05 models_derivatives_response_diversity.Rmd
run all the models to get the simplified and complex models, derivatives now

# predictions take VERY LONG! and two folders are needed to store the 
# DHARMa and prediction output inside: 
# total_models/gam_check
# total_models/predictions
# with three subfolders: temp, temp_test, depth_temp (total_models/predictions_depth_temp)

df_exclusion <- df_4 |> 
  filter(Species %in% c("Alburnus_arborella", "Barbatula_sp_Lineage_I",
                   "Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta", "Lepomis_gibbosus"))

# all temperature models, no simplificaiton

predictions(df_exclusion)

predictions(df_1)
predictions(df_2)
predictions(df_3)
predictions(df_4)

# temperature models with simplified models
predictions_temp(df_1)
predictions_temp(df_2)
predictions_temp(df_3)
predictions_temp(df_4)

# sensitivity
predictions_depth_temp(df_1)
predictions_depth_temp(df_2)
predictions_depth_temp(df_3)
predictions_depth_temp(df_4)

# issue with cottus sp po
# statistics of the gamms
total_models <- bind_rows(df_1, df_2, df_3, df_4)

stats_gamms(total_models)

#  compile data frame with all predictions from the saved species' model predictions

df_predictions_all <- list.files(path = "total_models/depth_temp/predictions", pattern = ".rds", full.names = TRUE) |>
  map_dfr(readRDS)

# save total predictions as RDS
saveRDS(df_predictions_all, "total_models/df_pred_all.rds")

df_predictions_all <- list.files(path = "total_models/predictions", pattern = ".rds", full.names = TRUE) |>
  map_dfr(readRDS)

@@

###AIC comparisons in special cases

AIC_testing <- function(data2, i){

  data <- data2 |> 
    filter(Species == i)

  data$fProtocol <- as.factor(data$Protocol)
  data$fLake <- as.factor(data$Lake)

  # model_0 <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) +
  #                  s(Depth_sample, k = 3)
  #                           + s(fLake, bs = 're')
  #                           +  s(fProtocol, bs = 're'), family = ziP())
  # print(glance(model_0))

model1 <- gam(data = data, Presence ~ s(mean_last_7days, k = 3) + s(Depth_sample, k = 3)
                            + s(fLake, bs = 're')
                            +  s(fProtocol, bs = 're'), family = binomial)
print(glance(model1))

model2 <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) +
                            + s(fLake, bs = 're')
                            +  s(fProtocol, bs = 're'), family = ziP())
print(glance(model2))

AIC_testing(df_4, "Lepomis_gibbosus")

# in Alosa agone: model_0 not running, smaller AIC for binomial depth + temp model -> will be modelled as binomial depth + temp
# same for Coregonus_sp_large_pelagic
# lepomis gibbosus also has the smallest AIC for binomial depth + temp model

# also testing the other special cases:
 # "Alburnus_arborella", "Barbatula_sp_Lineage_I","Cyprinus_carpio", "Phoxinus_csikii", "Salmo_trutta"

AIC_testing(df_4, "Alburnus_arborella") 
# not running with model0, Binomial with depth and temp best
AIC_testing(df_4, "Barbatula_sp_Lineage_I")
# binomial depth and temp has lowest AIC
AIC_testing(df_4, "Cyprinus_carpio")
# binomial depth and temp has lowest AIC 
AIC_testing(df_4, "Phoxinus_csikii")
#binomial depth and temp has lowest AIC 
AIC_testing(df_4, "Salmo_trutta")
# binomial depth and temp lowest AIC

# AIC_testing(df_4, "Cottus_sp_Po")
# # best model is the temp, depth model

AIC_testing(total_models, "Alburnus_arborella")
# 8 models are run with binomial temp and depth

```
####Fig. 2
Overview of all model predictions

@@

# separating the dfs to separate model types
depth_binomial_gam <- df_binomial_gam |> 
  filter(tot_abu > 20)

temp_binomial_gam <- df_binomial_gam |> 
  filter(tot_abu < 20)

depth_abundance_gam <- df_abundance_gam |> 
  filter(tot_abu > 20)

temp_abundance_gam <- df_abundance_gam |> 
  filter(tot_abu < 20)

depth_binomial_re <- df_binomial_re |> 
  filter(tot_abu > 20)

temp_binomial_re <- df_binomial_re |> 
  filter(tot_abu < 20)

depth_abundance_re <- df_abundance_re |> 
  filter(tot_abu > 20)

temp_abundance_re <- df_abundance_re |> 
  filter(tot_abu < 20)

@@

###Depth and temp models
```{r}
# DEPTH AND TEMPERATURE
# UNCLEAR HOW THIS CAN BE DONE
# Model 1: Binomial data without random intercepts

species_list <- depth_binomial_gam |>
  distinct(Species) |>
  pull(Species)

@@

i <- "Salmo_sp_Blackspot"
j <- "Poschiavo"
# loop through the list of species
for (i in species_list) {
  data <- depth_binomial_gam |>
    filter(Species == i)

@@

  gam_output[[i]] <- gam(data = data, Abundance ~ s(mean_last_7days, k = 3) +
                           s(Depth_sample, k = 3) + s(fProtocol, bs = 're'),
                         family = binomial)

@@

    # using minimum and maximum temp in the lake for derivatives
     newdata <- tibble(mean_last_7days = seq(from = min(data_lake$temp, na.rm = TRUE),
                                             to = max(data_lake$temp, na.rm = TRUE)
                                             ,length = 200), Depth_sample =
                         mean(data$Depth_sample, na.rm = TRUE),
                       fProtocol = factor("VERT"))

     # calculate derivatives
     # derivatives <- derivatives(gam_output[[i]], data = newdata) |>
     #   mutate(fLake = factor(j)) |>
     #   mutate(species = factor(i))
     #   rename(temp = data)
     #

       test_derivatives <- derivative_samples(gam_output[[i]]
                                              , focal = "mean_last_7days", data = newdata)

       test_derivatives |>
         filter(.draw <= 200) |>
         ggplot(aes(x = mean_last_7days, y = .derivative, group = .draw)) +
         geom_line(alpha = 0.5)

       derivatives <- test_derivatives |> 
         filter(.draw == 500) |> 
         mutate(fLake = factor(j)) |>
         mutate(species = factor(i))

       derivatives |> 
         ggplot(aes(mean_last_7days, .derivative)) +
         geom_line()

     saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
  }
}

# needs to be discussed

# nothing changed starting from here
############################################################################3##3
# Model 2.1: Abundance data and no random intercept for lake -> but binomial
# Coregonus_sp_benthic_profundal did not run with ZIP and was thus
# run with binomial.

species_list <- depth_abundance_gam |> 
  # binomial one
  filter(Species == "Coregonus_sp_benthic_profundal") |>
  distinct(Species) |> 
  pull(Species)

@@

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = data)

    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
  }
}

@@

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = data)

    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
  }
}

@@

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = data)

    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))

@@

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = data)
    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))

@@

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = data)

    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))

     # calculate derivatives
     derivatives <- derivatives(gam_output[[i]], data = newdata) |>
       mutate(fLake = factor(j)) |>
       mutate(species = factor(i)) |>
       rename(temp = data)

     saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
  }
}

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = data)

    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
  }
}

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = data)

    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))
  }
}

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = data)

    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = data)
    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = data)

    saveRDS(derivatives, paste0("total_models/derivatives/derivatives_", i, "_",  j, ".rds"))

>>>>

# predictions take VERY LONG! and two folders are needed to store the 
# DHARMa and prediction output inside: 
# total_models/gam_check
# total_models/predictions
# with three subfolders: temp, temp_test, depth_temp (total_models/predictions/depth_temp)

# all temperature models, no simplificaiton
# predictions(df_1)
# predictions(df_2)
# predictions(df_3)
# predictions(df_4)
# 
# no_exclusion_predictions_all <- list.files(path = "total_models/predictions/temp_test", pattern = ".rds", full.names = TRUE) |>
#   map_dfr(readRDS)
# 
# # save total predictions as RDS
# saveRDS(no_exclusion_predictions_all, "total_models/df_pred_no_exclusion.rds")

# temperature models with simplified models

predictions_temp(df_1)
predictions_temp(df_2)
predictions_temp(df_3)
predictions_temp(df_4)

no_exclusion_predictions_all <- list.files(path = "total_models/predictions/temp", pattern = ".rds", full.names = TRUE) |>
  map_dfr(readRDS)

# save total predictions as RDS
saveRDS(no_exclusion_predictions_all, "total_models/df_pred_all.rds")

# sensitivity
# predictions_depth_temp(df_1)
# predictions_depth_temp(df_2)
# predictions_depth_temp(df_3)
# predictions_depth_temp(df_4)
# 
# # issue with cottus sp po
# # statistics of the gamms
# total_models <- bind_rows(df_1, df_2, df_3, df_4)
# 
# stats_gamms(total_models)
# 
# #  compile data frame with all predictions from the saved species' model predictions
# 
# depth_temp_predictions_all <- list.files(path = "total_models/predictions/depth_temp", pattern = ".rds", full.names = TRUE) |>
#   map_dfr(readRDS)
# 
# # save total predictions as RDS
# saveRDS(depth_temp_predictions_all, "total_models/depth_temp_pred_all.rds")

@@

####Fig. 2
Overview of all model predictions

@@

@@

@@

@@

@@

     # calculate derivatives
     derivatives <- derivatives(gam_output[[i]], data = newdata) |>
       mutate(fLake = factor(j)) |>
       mutate(species = factor(i)) |>
       rename(temp = mean_last_7days)

     saveRDS(derivatives, paste0("total_models/derivatives/simple/derivatives_", i, "_",  j, ".rds"))
  }
}

@@

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i))
      rename(temp = mean_last_7days)

    saveRDS(derivatives, paste0("total_models/derivatives/simple/derivatives_", i, "_",  j, ".rds"))
  }
}

@@

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = mean_last_7days)

    saveRDS(derivatives, paste0("total_models/derivatives/simple/derivatives_", i, "_",  j, ".rds"))
  }
}

@@

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = mean_last_7days)

    saveRDS(derivatives, paste0("total_models/derivatives/simple/derivatives_", i, "_",  j, ".rds"))

@@

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = mean_last_7days)
    saveRDS(derivatives, paste0("total_models/derivatives/simple/derivatives_", i, "_",  j, ".rds"))

@@

    derivatives <- derivatives(gam_output[[i]], data = newdata) |>
      mutate(fLake = factor(j)) |>
      mutate(species = factor(i)) |>
      rename(temp = mean_last_7days)

    saveRDS(derivatives, paste0("total_models/derivatives/simple/derivatives_", i, "_",  j, ".rds"))
